---
title: Competition
author: 
  - name: Hailey Cantwell
    email: hailey.cantwell@snhu.edu
    affiliations: 
      - name: Southern New Hampshire University
format: html
date: 1/8/2024
date-modified: today
date-format: long
theme: night
toc: true
code-fold: true
---

#### Loading Data
```{r}
#| echo: false
#| message: false
#| warning: false
library(tidyverse)
library(tidymodels)
library(kableExtra)
library(patchwork)
library(ggridges)
library(marginaleffects)
library(tidytext)

options(kable_styling_bootstrap_options = c("hover", "striped"))

theme_set(theme_bw(base_size = 14))

data <- read_csv("data.csv") %>%
  janitor::clean_names()
comp <- read_csv("comp.csv") %>%
  janitor::clean_names()
data <- data %>%
  select(-c(description, city, id))

set.seed(123)
data_splits <- initial_split(data, prop = 0.85, strata = price_range)
train <- training(data_splits)
temp <- testing(data_splits)

set.seed(456)
test_splits <- initial_split(temp, prop = .6)
validation <- training(test_splits)
test <- testing(test_splits)
```

## Statement of Purpose
The purpose of this project is to develop a machine learning model that can predict real estate property price ranges of houses in Austin, TX. This will be utilized by real estate agencies and property buyers/sellers. Accurate predictions may assist in making informed decisions regarding property value on Zillow. 

## Exploratory Data Analysis

```{r}
head(data) %>%
  kable() %>%
  kable_styling()
```
```{r}
train %>%
  count(price_range) %>%
  arrange(price_range)
```

```{r}
train %>%
  select(where(is.numeric)) %>%
  summary() %>%
  kable() %>%
  kable_styling(full_width = FALSE, bootstrap_options = "striped")
```
## Distribution of Price Range

```{r}
train <- train %>%
  mutate(estimated_price = case_when(
    price_range == "0-250000" ~ 125000,
    price_range == "250000-350000" ~ 300000,
    price_range == "350000-450000" ~ 400000,
    price_range == "450000-650000" ~ 550000,
    price_range == "650000+" ~ 850000,
    TRUE ~ NA_real_
  ))


```

```{r}
train_clean <- train %>%
  filter(!is.na(estimated_price))


```

```{r}
train %>%
  summarize(
    min_price = min(estimated_price, na.rm = TRUE),
    median_price = median(estimated_price, na.rm = TRUE),
    avg_price = mean(estimated_price, na.rm = TRUE),
    max_price = max(estimated_price, na.rm = TRUE),
    sd_price = sd(estimated_price, na.rm = TRUE)
  ) %>%
  pivot_longer(everything(),
               names_to = "Metric",
               values_to = "Value") %>%
  kable(digits = 0) %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("hover", "striped"))

```
### Estimate Price Distribution 
```{r}
train_clean %>%
  ggplot() + 
  geom_histogram(aes(x = estimated_price), 
                 fill = "purple", 
                 color = "black", 
                 bins = 5) +
  scale_x_continuous(labels = scales::label_dollar()) +
  labs(
    title = "Estimated Price Distribution",
    x = "Estimated Price",
    y = "Count"
  ) +
  theme_minimal()

    
```

### Project Price Range by Bedrooms
```{r}
train_clean %>%
  ggplot(aes(x = as.factor(num_of_bedrooms), y = estimated_price)) +
  geom_boxplot(fill = "skyblue") +
  scale_y_continuous(labels = scales::label_dollar()) +
  labs(
    title = "Estimated Price by Number of Bedrooms",
    x = "Number of Bedrooms",
    y = "Estimated Price"
  ) +
  theme_minimal()


```
